Optimal design of experiment

### 2025-01-11 Article_Design_of_Experiment-Chamomile.tex
30.10
11.01

\subsection{Optimal design of experiment} \label{CH: DOE}
\subfile{Sections/Optimal_design_of_experiment}

@@

\section{Conclusions} \label{CH: Conclusion}
This paper introduces the D-optimal experimental design (D-OED) as a model-based method for finding an experimental procedure in chemical engineering problems. D-OED forms the objective function, which can be interpreted as minimising the volume of uncertainty ellipsoid. Compared to the classical formulation of D-OED, a penalty term was introduced to the cost function to prevent infeasibly rapid changes in operating conditions or bang-bang type control. The method is demonstrated through a case study on supercritical extraction. This work aims to design an experiment to improve the precision of the correlation for $D_i$. While the original experiments were conducted under constant operating conditions, this work looks for an experimental procedure with dynamically changing operating conditions. In such a case, the D-OED solution can also be used for model validation.

The analysis was repeated for multiple pressure values, and the obtained optimal profiles of inlet temperature and mass flow rate were similar across all cases. The resulting yield curves exhibit wavy patterns due to the relationship between the Hessian and the curvature of a multi-variable function. The analysis of the yield curves and scatter plots reveals that applying the same experimental strategy at different operating conditions results in varying amounts of information being collected. Consequently, it is concluded that the operating conditions should always be carefully chosen to conduct the most informative experiments. This is because the physical properties of CO$_2$ vary less at pressures far from the critical point, meaning that the same set of controls has less impact on the system at high pressures than at low pressures.

% ===================================================
% Bibliography
% ===================================================
%% Loading bibliography style file
\newpage
%\bibliographystyle{model1-num-names}
\bibliographystyle{unsrtnat}
\bibliography{mybibfile}

@@

\newpage

>>>>

\subsection{Optimal design of experiment} \label{CH: DOE}
\subfile{Sections/DOE}

@@

\section{Conclusions} \label{CH: Conclusion}
This paper introduce the D-optimal experimental design (D-OED) as a model-based method for determining optimal experimental procedures in chemical engineering problems. D-OED optimizes an objective function, which can be geometrically interpreted as minimizing the volume of the uncertainty ellipsoid, thereby maximizing the precision of parameter estimates. Compared to the classical formulation of D-OED, this work introduces a penalty term to the cost function to prevent unrealistic rapid changes in operating conditions, ensuring practical feasibility in experimental implementation. The method is demonstrated through a case study on supercritical extraction, focusing on designing experiments to improve the precision of the correlation for the diffusivity coefficient $D_i$. While the original experiments were conducted under constant operating conditions, this study explores dynamically changing operating conditions, highlighting the potential of D-OED solutions for both parameter estimation and model validation.

The analysis was conducted for multiple pressure cases, and the optimal mass flow rate profiles were similar across all instances. The inlet temperature profiles show different patterns, but all of them start with high temperature and decrease in the second half of the batch. The resulting yield curves exhibit distinct wavy patterns, which can be explained through the relationship between the Hessian matrix and the Gaussian curvature of a multi-variable function. The determinant of the Fisher information matrix, central to the D-optimality criterion, is proportional to the Gaussian curvature of the parameter manifold. This curvature reflects the system's sensitivity to parameter changes and is maximized by the D-OED method. The observed wavy behavior of the yield curves is thus a manifestation of the high-curvature regions identified by the optimization, where the system's response to experimental conditions is most informative.

Further analysis of the yield curves and scatter plots reveals that the informativeness of experiments varies significantly across different operating conditions. This variation depends on the physical properties of CO$_2$, which strongly change around the critical point. Consequently, it is concluded that operating conditions must be carefully selected to the specific regime of interest to maximize the information gained. This observation emphasizes the importance of aligning experimental strategies with the system's underlying physical properties and model structure.

% ===================================================
% Bibliography
% ===================================================
%% Loading bibliography style file
\clearpage
\newpage
%\bibliographystyle{model1-num-names}
\bibliographystyle{unsrtnat}
\bibliography{mybibfile}

@@

\clearpage
\newpage

### 2025-01-11 Sections/Results_DOE.tex
11.01

	\label{CH: Results}

	To identify the global solution of Equations \ref{EQ:Formulation_1}, the optimization problem is solved multiple times, each starting from a random initial solution. Figure \ref{fig:scatter_T} compares the initial and final values of the cost function across multiple optimization runs. The shaded area spans from the smallest to the third smallest final values of the objective function. The black curves on the left and below the scatter plot represent the distributions of the initial and final cost function values, respectively. From Figure \ref{fig:scatter_T}, it can be observed that multiple local solutions exist for Case 1. The solution with the lowest cost function value is considered the global solution.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/scatter_T.png}	
		\caption{Initial vs final values of the cost function}
		\label{fig:scatter_T}
	\end{figure}

	The optimal profiles of inlet temperature and mass flow rate are presented in Figure \ref{fig:profile_T}. One of the independent parameters in the analysed correlation is the mass flow rate, which is directly modified several times during the experiment. The second independent parameter is the Reynolds number, which can be affected by changes in the thermodynamic properties of the fluid. The thermal properties of CO$_2$ between $30^\circ$C and $40^\circ$C, assuming a constant pressure of 150 bar, vary only slightly. For instance, the dynamic viscosity changes from 67 to $80 \times 10^{-6}$ Pa$\cdot$s. The obtained profile of the inlet temperature shows minimal variation, suggesting a limited influence of inlet temperature on the extraction yield. Given the narrow operational window for inlet temperature at constant pressure, the mass flow rate becomes the primary variable to control.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/Profile_T_1.png}	
		\caption{Optimal temperature and mass flow rate profiles}
		\label{fig:profile_T}
	\end{figure}

	Similar to Case 1, Case 2 is solved multiple times starting from random initial solutions to identify the global solution. Figure \ref{fig:scatter_P} summarizes the changes in the objective function values across all optimization runs. It is evident that both the initial and final values of the cost function are significantly lower than any values obtained in Case 1. In Case 2, the wider operating range leads to more significant changes in the physical properties of CO$_2$ compared to Case 1. For example, the kinematic viscosity of CO$_2$ ranges from $48 \cdot 10^{-6}$ (at $40^\circ$C and 100 bar) to $90 \cdot 10^{-6}$ Pa$\cdot$s (at $30^\circ$C and 200 bar). The greater variability in pressure induces a more complex system response, including temperature variations, resulting in lower values of the objective function compared to Case 1.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/scatter_P.png}	
		\caption{Initial vs final values of the cost function}
		\label{fig:scatter_P}
	\end{figure}

	The optimal profiles of pressure and mass flow rate are presented on Figure \ref{fig:profile_P}. Similarly to case 1, only few changes in the mass flow rate are observed in the plot. In contrary to the inlet temperature, the pressure changes are frequently applied in the case 2.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/Profile_P_1.png}	
		\caption{Optimal pressure and mass flow rate profiles}
		\label{fig:profile_P}
	\end{figure}

>>>>

	To identify the global solution for Equations \ref{EQ:Formulation_2}, the optimization problem is solved multiple times, each run starting from a random initial solution sampled from a uniform distribution. Figure \ref{fig:scatter} compares the initial and final cost function values across multiple optimization runs for all pressure cases. The solution with the lowest cost function value is considered the global solution for each case.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/scatter.png}	
		\caption{Initial vs final values of the cost function}
		\label{fig:scatter}
	\end{figure}

	From Figure \ref{fig:scatter}, it is evident that multiple local solutions exist. By analyzing locations of the clusters in Figure \ref{fig:scatter}, it can be concluded that experiments conducted near the critical point provide more information regarding the correlation than those conducted farther from it. For each pressure case, the profiles with the lowest value of the objective function are further analysed. The closer the pressure is to the critical point, the larger the deviations in the physical properties of $CO_2$ caused by changes in the controls, leading to greater variation in the Reynolds number, and consequently to more informative experiments.

	As presented in Figure \ref{fig:profiles_F}, all mass flow rate profiles initially stay at a minimum value, persisting for 100 to 150 minutes, depending on the pressure. Subsequently, a step-like increase in mass flow rate is observed, with the onset occurring earlier at higher pressures. Between 250 and 300 min, the mass flow rate stabilizes at value of the upper bound.

	\begin{figure}[t!]
		\centering
		\begin{subfigure}[t]{\columnwidth}
			\centering
			\includegraphics[width=\columnwidth]{Figures/Results/Profile_F.png}	
			\caption{Optimal mass flow rate profiles}
			\label{fig:profiles_F}
		\end{subfigure}%
		\par\bigskip % force a bit of vertical whitespace
		\begin{subfigure}[t]{\columnwidth}
			\centering
			\includegraphics[width=\columnwidth]{Figures/Results/Profile_T.png}	
			\caption{Optimal inlet temperature profile}
			\label{fig:profiles_T}
		\end{subfigure}
		\par\bigskip % force a bit of vertical whitespace
		\begin{subfigure}[t]{\columnwidth}
			\centering
			\includegraphics[width=\columnwidth]{Figures/Results/yield.png}	
			\caption{Optimal yield profiles}
			\label{fig:profiles_y}
		\end{subfigure}
		\caption{Results of the optimization problem}
	\end{figure}

	Figure \ref{fig:profiles_T} shows the optimal profiles of the inlet temperature ($T^{in}$) for various pressure cases in the supercritical CO$_2$ extraction process. Initially, the $T^{in}$ remains constant at approximately $40~^\circ C$ for all pressures, indicating a stable starting condition for the system. The temperature decreases after about 300 min, but the exact time slightly varies depending on the pressure. This decline is rapid for lower pressures and gradual for higher pressures. The blue curve (100 bar) exhibits the sharpest temperature drop, while the green curve (200 bar) demonstrates a slower and more gradual decline. At lower pressures, the final temperature reaches the lower bound and remains steady until the end of the batch. In contrary, the higher pressure cases result in a less significant temperature reduction. Moreover, the system does not reach steady operating conditions, as the inlet temperature profile does not flatten by the end of the process.

	Figure \ref{fig:profiles_y} depicts the predicted yield curves. A closer examination of the yield and scatter plots reveals that the most informative experiments do not necessarily correspond to those with the highest yield. When the D-optimality criterion is applied, the aim is to maximize the determinant of the Fisher information matrix across the experimental design space. This approach identifies conditions that amplify variations in observed responses, particularly in regions with high sensitivity to parameter changes. As a result, the yield curve may display a "wavy" pattern, characterized by oscillations rather than a smooth, monotonic trend. These fluctuations arise from increased multiple critical points—local maxima, minima, or saddle points—on the response surface. By maximizing the sum of determinants of the Fisher information matrix, the experimental design prioritizes regions of high curvature, where the response changes sharply.

	Alternatively, the D-optimality problem can be understood through the concept of Gaussian curvature. In this view, the experimental design shapes the parameter space of a model into a manifold with local geometry defined by the Fisher information matrix. The D-optimality condition maximizes the determinant of this matrix, which geometrically corresponds to maximizing the product of the principal curvatures of the manifold at a given point. This determinant is directly proportional to the Gaussian curvature, linking experimental design to the geometry of the parameter space. The determinant of the Fisher information matrix can be interpreted as a measure of the "volume element" on the parameter manifold, with higher values indicating regions of greater local curvature. By maximizing this determinant, the parameter manifold is locally as curved as possible, which explains the wavy behavior of the yield curves.

	The results and observations discussed in this article are local and should be considered valid only for the model given by \citet{Sliczniuk2024}. Although general conclusions about the curvature of the yield curve are independent of a process model, a completely different set of control profiles should be expected if the same technique is applied to a different model. Such strong dependence on a model is one of the most significant limitations of this work. Moreover, the model predicts the dataset in the region, which has not been validated during previous experiments. One hand this leaves a room for model validation, on the other hand the obtained set of data points might differs from the predictions. In such a case, further improvement of the model is necessary.
